When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents

### 2024-10-10 article_doc.Rmd
format changes JPCS

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"
author:
  "Ismail Guennouni^1,2,3,4^*, Georgia Koppe^1,2,3^$^\\dagger$, Christoph Korn^4^$^\\dagger$"
abstract: |
bibliography: "bib/exposure.bib"
csl: apa.csl
output:
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
header-includes:
  - \usepackage{lscape}
editor_options: 
  markdown: 
    wrap: 72
---

\small

^1^ *Department of Psychiatry and Psychotherapy, Central Institute of
Mental Health, Medical Faculty Mannheim, Heidelberg University,
Mannheim, Germany*

@@

^\*^ *Corresponding author. Address: Central institute of Mental Health,
J5, Mannheim, Germany. Email:
[ismail.guennouni\@zi-mannheim.de](mailto:ismail.guennouni@zi-mannheim.de){.email}*

**Abstract:**

Can exposure to forgiving behavior improve cooperative outcomes, and
does Rejection Sensitivity (RS) moderate this effect? This randomized
controlled experiment explores these questions using a novel approach
with Hidden Markov Model (HMM)-based artificial agents in repeated trust
games (RTGs). Participants (N = 206), pre-screened for high or low RS,
interacted with either consistently behaving or more forgiving
co-players, which were in reality HMM agents. Unexpectedly, exposure to
forgiving agents led to reduced cooperation in subsequent interactions.
Control group participants maintained consistent behavior. RS levels
influenced perceptions of the agent's cooperativeness but did not
significantly affect behavioral outcomes, revealing a
perception-behavior dissociation. These findings challenge assumptions
about fostering cooperation through simple exposure to positive
interactions and highlight the importance of accounting for negative
contrast effects when designing such interventions. They also
demonstrate the potential of HMM-based artificial agents for studying
interactive social dynamics, offering a methodological advancement for
future research. Our results suggest the need for nuanced interventions
that gradually shape expectations and behaviors to foster cooperative
outcomes, particularly addressing the discrepancy between perceptions
and behavior in individuals with high rejection sensitivity.

@@

**Keywords** : Interpersonal functioning; Rejection Sensitivity;
Forgiveness Intervention; Trust-based Cooperation; Hidden Markov Models

@@

**General Scientific Summary:**

This study reveals that exposing individuals to forgiving behavior in
economic games can unexpectedly decrease their subsequent cooperation.
Interestingly, while people with high rejection sensitivity more
accurately perceived their partners' cooperativeness, this didn't
translate into more cooperative behavior. These findings challenge
assumptions about fostering cooperation and highlight the need for more
nuanced interventions to promote reciprocity, especially for individuals
prone to expecting rejection.

@@

One explanation for such social challenges lies in early caregiver
experiences. Attachment theory [@bowlby_attachment_1978] suggests that
the quality of these relationships shapes our capacity for secure
attachments and trust. Individuals with higher levels of insecure
attachment may recall negative trust-related experiences more easily,
report fewer positive trust experiences, and use less constructive
coping strategies when trust is broken [@mikulincer_attachment_1998].
These insecure attachment patterns are often associated with heightened
rejection sensitivity (RS), a tendency to anxiously expect, readily
perceive, and intensely react to rejection [@downey_implications_1996;
@downey_early_1997]. RS has been linked to the development of various
mental health conditions, including depression, anxiety, personality
disorders, and self-harm [@gao_associations_2017]. Individuals high in
RS show attentional biases towards social threat cues, which may
contribute to difficulties in social interactions
[@berenson_rejection_2009]. A recent meta-analysis revealed prosocial
behavior and interpersonal trust as two key processes of interpersonal
functioning that are markedly impaired in PDs and which are likely to
contribute to interpersonal dysfunction in this population
[@hepp_prosociality_2022]. The interaction of RS and trust-based
constructs has been explored, particularly in Borderline Personality
Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found
that RS mediated the relationship between BPD features and lower trust
appraisal. @abramov_influence_2022-1 found that higher baseline feelings
of rejection in individuals with BPD predict slower trust formation and
less pronounced declines in trust following trust violations during the
trust game. However, the interaction between *reciprocity* and RS hasn't
been studied as extensively, leaving a gap in our understanding of how
these constructs might interplay. Could an ingrained sensitivity to
rejection contribute to reduced pro-social behavior such as a failure to
show reciprocity as a result of perceived rejection by the interaction
partner?

Given that RS may be a manifestation of maladaptive attachment styles,
it's crucial to explore whether exposure to consistently forgiving
interaction partners could help reshape these interpersonal expectations
and behaviors. If RS indeed influences reciprocity in repeated
interactions, could such exposure serve as a corrective experience,
potentially improving interpersonal outcomes? This approach aligns with
attachment theory, which suggests that new, positive relational
experiences can modify internal working models of relationships
[@bowlby_secure_1988]. Research in the fields of behavioral economics
and psychology has also explored how positive social interactions
influence cooperation and pro-social behavior. The use of the repeated
trust game (RTG), a well-established experimental approach, has allowed
for the analysis of the development of trust through ongoing
interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual
trust, where each party's trust is reciprocated with trustworthiness,
have the effect of enhancing cooperative behaviors even among
individuals who are initially inclined to be distrustful
[@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in
social networks interacting in a public goods game and found that
cooperative behavior tends to cluster, suggesting that exposure to
cooperative peers can lead to more cooperative behavior. Similarly,
research on social learning theory [@bandura_social_1977] has long
demonstrated that individuals learn and model the behavior of those
around them, indicating that if someone is consistently exposed to
cooperative and positive individuals, they're likely to emulate this
behavior. These insights highlight that engaging with compassionate and
forgiving others can be an effective method for mitigating deeply
ingrained mistrust and potentially fostering trustworthy behavior.

In this study, we use a randomized controlled online experiment to test
whether exposing participants with varying RS levels to forgiving and
more cooperative co-players results in more trustworthy behavior and a
repair of potential breakdowns in RTG cooperation. To simulate realistic
social interaction while maintaining a high degree of experimental
control, we take a novel paradigmatic approach: We use generative models
of how humans play the RTG to design an agent that plays the role of the
investor, based on Hidden Markov Models (HMMs) fitted to real players'
data. A key aspect of these agents is that their actions depend on a
latent "trust state" which reacts dynamically to the trustees' returns,
simulating real-life trust-building scenarios. An advantage of having
such a generative model of behavior is the possibility of controlling
different aspects of the agent's strategy such as its general policy,
the propensity to cooperate actively, or the propensity to trust again
after breakdowns of cooperation. To further mimic real-world
interactions and examine participants' responses to one-off breakdowns
of cooperation, we incorporate occasional pre-programmed low investments
by the agent.

We pre-screened participants for high or low RS using a validated
questionnaire, then assigned them exclusively to the trustee role in a
series of trust games. After playing a 15-round RTG with a human-like
HMM investor, they were randomly assigned to either a Control or
Manipulation condition. In the Manipulation condition, participants were
exposed over three RTGs to HMM investors designed with a limited
propensity for retaliation, potentially mitigating ingrained mistrust.
In the Control condition, participants played three RTGs against the
same human-like HMM. After this exposure phase, all participants played
another 15-round RTG with a human-like HMM investor, similar to the one
in the pre-exposure phase. We expect those exhibiting high RS to be more
responsive to the pre-programmed change in the agent's cooperativeness
and retaliation propensity. We also hypothesize that those in the
Manipulation condition would behave more cooperatively and have a lower
propensity to retaliate to the co-player's defection after the exposure
phase, with those in the higher RS group showing a bigger effect.

@@

To have participants with large differences in RS, a total of 1195
participants were pre-screened on the Prolific Academic platform
(prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to
finally select two similarly sized groups: One with high RS (RSQ score
\> 15, N=103) and the other with low RS (RSQ score \< 10, N=103)
totalling 206 participants (56% female). These were then invited through
prolific to take part in the main experiment. The required sample size
was determined using an *a priori* power analysis to have an 80%
probability to detect a small effect size (Cohen’s f = 0.10) for a
within-between interaction with a 5% type I error rate in a repeated
measures ANOVA. The sample size calculation assumed 2 groups, 2
measurement per group and was performed using the G\*Power software
[@faul_statistical_2009]. The mean age of participants was 34.6 years,
with an 11.9 years standard deviation. The majority of participants
identified ethnically as White ($80$%). The online cohort registered 30
unique countries of birth with the most frequent being the U.K ($33$%)
followed by Poland ($10$%) and Portugal ($10$%). Participants were paid
a fixed fee of £6 plus a bonus payment dependent on their performance
that averaged £0.5. Data was collected over multiple sessions between
December 2023 and February 2024.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS :
High or Low) by 2 (Phase: Trust-Game Pre Manipulation, Trust-Game Post
Manipulation) design, with repeated measures on the Phase factor (Figure
\@ref(fig:HMMPanels).A). Participants within each pre-screened RS group
were randomly assigned to one of the two levels of the Condition factor,
resulting in 101 participants in the Manipulation condition and 105 in
the Control condition. The games were designed and implemented online
using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment
received approval from the University of Heidelberg's Medical Faculty
ethics commission (ID:S-708/2023) and the experiment was performed in
accordance with the ethics board guidelines and regulations. All
participants provided informed consent prior to their participation.

@@

````{=tex}
\begin{landscape}
```{r HMMPanels, fig.cap= "\\small{A: A timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.}",include=T, out.width='1\\linewidth', out.height='1\\textheight', fig.align='center'}

@@

# Determine the target width for all images, for example, the width of the widest image
target_width <- max(image_info(image1)$width, image_info(image2)$width, 
                    image_info(image3)$width, image_info(image4)$width)

target_height <- max(image_info(image1)$height, image_info(image2)$height, 
                    image_info(image3)$height, image_info(image4)$height)

# Scale images to have the same width
image1 <- image_scale(image1, paste0(target_width, "x", target_height))
image2 <- image_scale(image2, paste0(target_width, "x", target_height))
image3 <- image_scale(image3, paste0(target_width, "x", target_height))
image4 <- image_scale(image4, paste0(target_width, "x", target_height))

# Annotate each image with a label
image1 <- image_annotate(image1, text = "A", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image2 <- image_annotate(image2, text = "B", location = "+0+10", size = 90, color = "black", gravity = "northwest")
image3 <- image_annotate(image3, text = "C", location = "-0+10", size = 90, color = "black", gravity = "northwest")
image4 <- image_annotate(image4, text = "D", location = "+0+10", size = 90, color = "black", gravity = "northwest")

# Assuming a 300 DPI resolution for good print quality
dpi <- 300

# Spacing of 0.1 inches converted to pixels
spacing <- 0.05 * dpi

# Calculating the width and height for each image
width_per_image <- ((12 - 0.1) / 2) * dpi
height_per_image <- ((7.4 - 0.1) / 2) * dpi
# The geometry string for each image, including spacing (assuming no borders)
geometry_string <- paste0(width_per_image, 'x', height_per_image, '+', spacing, '+', spacing)

# Now create the montage with the specified geometry
combined_image <- image_montage(
  c(image1, image2, image3, image4), 
  tile = "2x2",
  geometry = geometry_string
)

# Display the combined image
plot(combined_image)
```
\end{landscape}
````

@@

```{r}
# Number of participants in manipulation condition
temp_manip <- avg_ret_df %>% filter(condition.f =="Manipulation")

length(unique(temp_manip$playerId))

# Number of participants with high RS

temp_RS <- avg_ret_df %>% filter(high_RS =="high RS")
length(unique(temp_RS$playerId))

@@

<!-- EFFECT SIZES -->

```{r}
# Load the necessary packages

@@

To examine this four-way interaction, we conducted a contrast analysis
of how the effect of investment on returns changed from pre- to
post-manipulation for different RS groups in both conditions. Starting
with the Manipulation condition, for participants with low RS, the
effect of investment on returns increased significantly from pre- to
post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipLowRS`.
This suggests that after the manipulation, low RS participants became
more responsive to their co-player's investments, returning
proportionally more as investments increased. In contrast, for
participants with high RS, the effect of investment on returns decreased
significantly from pre- to post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipHighRS`.
This indicates that high RS participants became less responsive to their
co-player's investments after the manipulation, with smaller increases
in returns as investments increased. The difference in these pre-post
changes between high and low RS groups was significant,
`r papaja::apa_print(results_4w)$full_result$DiffinPost_PrebetweenHighandLowRS`.
This result suggests that the manipulation had significantly different
effects on how low and high RS participants responded to their
co-player's investments.

@@

<!-- Focusing on post defection only:  -->

### Post Defection Trials

@@

### HMM Investor in Pre and Post Phases

<!-- Linear mixed effects model for investments -->

Was the HMM's strategy similar between pre and post phases in the
control condition? Was participants' behavior post exposure
differentiated enough to induce a different reaction from the HMM? To
answer these questions, we test for differences in the HMM agent's
investment by Phase, Condition and RS using a linear mixed-effects model
as described in the methods section. As seen in Figure
\@ref(fig:boxPlots), we find no main or interaction effects, indicating
the HMM's behavior was on aggregate similar across levels of Phase,
Condition and RS. This consistency in the investor's behavior is a
desirable feature of the HMM agent when the participants behavior is
largely simialr between phases. More importantly, it indicates that the
lower returns of participants in the post phase of the manipulation
condition were not differentiated enough to make the HMM react by
transitioning to lower latent trust states. It is also noteworthy that
the HMM agent was relatively successful in imitating human behavior in
this paradigm: When asked during debrief whether they thought the
investors they faced were human or not, $41$% of participants thought
they were either facing a human or were not sure of the nature of the
co-player. When asked to justify their choice, many answers reflected
participants projecting human traits such as "spitefulness" or "greed"
onto the artificial co-player's behavior.

@@

emcheck_inv <- pairs(emmeans::emmeans(mod_invs_expo, c("condition.f"), pbkrtest.limit = 6300))
emcheck_inv
```

So far we focused on analysing behavior for the pre and post phases.
Here, we look at returns and investments in the exposure phase. The
linear mixed effects model of participants' returns in the exposure
phase does not show a main effect of Condition on returns. There was a
main effect of Investment,
`r papaja::apa_print(mod_returns_expo)$full_result$inv_scaled`, with
participants positively reciprocating higher investments, an interaction
effect between Condition and Investment
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled)`,
showing a stronger positive reciprocity in the Control condition, and
finally a three way interaction between the RS group, Condition and
Investment
`r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled_high_RS)`,
showing that this stronger positive reciprocity to investment in the
Control condition is higher for participants with high RS. The linear
mixed effects model of the HMM investments shows a main effect of
Condition `r papaja::apa_print(mod_invs_expo)$full_result$condition_f`,
suggesting higher overall investments for the forgiving HMM compared to
the human-like HMM, but no difference in investments when facing low and
high RS groups.

@@

### Comparing Pre and Post Ratings

<!-- Comparing post-exposure to exposure ratings, participants in the Manipulation condition rated the investors they faced in the exposure phase as less cooperative,  -->

<!-- `r papaja::apa_print(comp3_coop)$full_result$Manipulation_PostvsExpo`, less forgiving `r papaja::apa_print(comp3_forg)$full_result$Manipulation_PostvsExpo`, and were less willing to face them again `r papaja::apa_print(comp3_again)$full_result$Manipulation_PostvsExpo`. Ratings for those in the Control group did not differ on how cooperative, forgiving the investors were, and on willingness to face them again.  -->

@@

```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

@@

We used a randomized controlled online experiment where participants
played a RTG with artificial agents designed to simulate human-like
trust-building scenarios. Participants were then exposed to either
forgiving or human-like HMM agents before playing another RTG. We found
that RS did not moderate participants' returns as trustees in the
repeated trust game. While previous research has shown that RS affects
*trust* formation, appraisal and repair, its impact on *reciprocity* in
repeated economic exchanges has been less explored. Our results suggest
a potential dissociation between RS's known effects on broader social
behavior and its limited influence on reciprocity in structured,
repeated interactions, challenging assumptions about the pervasive
influence of RS on social behavior and highlighting the complexity of
factors influencing reciprocity in economic exchanges.

>>>>

---
title: "When Forgiveness Backfires: Rejection Sensitivity and Cooperative Behavior Following Exposure to Adaptive Forgiving Agents"

author:
  "Ismail Guennouni^1,2,3,4^*, Georgia Koppe^1,2,3^$^\\dagger$, Christoph Korn^4^$^\\dagger$"

abstract: |
  Can exposure to forgiving behavior improve cooperative outcomes, and does Rejection Sensitivity (RS) moderate this effect?  This randomized controlled experiment explores these questions using a novel approach with Hidden Markov Model (HMM)-based artificial agents in repeated trust games (RTGs). Participants (N = 206), pre-screened for high or low RS, interacted with either consistently behaving or more forgiving co-players, which were in reality HMM agents. Unexpectedly, exposure to forgiving agents led to reduced cooperation in subsequent interactions. Control group participants maintained consistent behavior. RS levels influenced perceptions of the agent's cooperativeness but did not significantly affect behavioral outcomes, revealing a perception-behavior dissociation. These findings challenge assumptions about fostering cooperation through simple exposure to positive interactions and highlight the importance of accounting for negative contrast effects when designing such interventions. They also demonstrate the potential of HMM-based artificial agents for studying interactive social dynamics, offering a methodological advancement for future research. Our results suggest the need for nuanced interventions that gradually shape expectations and behaviors to foster cooperative outcomes, particularly for individuals with high rejection sensitivity.

bibliography: "bib/exposure.bib"
csl: apa.csl

output:
  bookdown::word_document2:
    fig_caption: true
    number_sections: false 
---
\small
^1^ *Department of Psychiatry and Psychotherapy, Central Institute of Mental Health, Medical Faculty Mannheim, Heidelberg University, Mannheim, Germany*

@@

^\*^ *Corresponding author. Address: Central institute of Mental Health, J5, Mannheim, Germany. Email: ismail.guennouni@zi-mannheim.de*

@@

**Keywords** : Social Trust; Forgiveness Intervention; Rejection Sensitivity; Interpersonal Dynamics; Hidden Markov Models

@@

**Highlights:**

- Exposure to forgiving artificial agents unexpectedly led to reduced cooperation in subsequent interactions, challenging assumptions about fostering trust through positive experiences

- Rejection sensitivity influenced perceptions of artificial agents' cooperativeness but not behavioral outcomes, revealing a dissociation between perception and behavior

- Novel Hidden Markov Model-based artificial agents provide adaptive, human-like behavior in trust games, allowing for precise control of "cooperative" and "forgiving" strategies while maintaining experimental control of co-player strategies

@@

One explanation for such social challenges lies in early caregiver experiences. Attachment theory [@bowlby_attachment_1978] suggests that the quality of these relationships shapes our capacity for secure attachments and trust. Individuals with higher levels of insecure attachment may recall negative trust-related experiences more easily, report fewer positive trust experiences, and use less constructive coping strategies when trust is broken [@mikulincer_attachment_1998]. These insecure attachment patterns are often associated with heightened rejection sensitivity (RS), a tendency to anxiously expect, readily perceive, and intensely react to rejection [@downey_implications_1996; @downey_early_1997]. RS has been linked to the development of various mental health conditions, including depression, anxiety, personality disorders, and self-harm [@gao_associations_2017]. A recent meta-analysis revealed prosocial behavior and interpersonal trust as two key processes of interpersonal functioning that are markedly impaired in PDs and which are likely to contribute to interpersonal dysfunction in this population [@hepp_prosociality_2022]. The interaction of rejection sensitivity and trust-based constructs has been explored, particularly in Borderline Personality Disorder (BPD). @miano_rejection_2013 and @richetin_emotional_2018 found that rejection sensitivity mediated the relationship between BPD features and lower trust appraisal. @abramov_influence_2022-1 found that higher baseline feelings of rejection in individuals with BPD predict slower trust formation and less pronounced declines in trust following trust violations during the trust game. However, the interaction between trustworthiness and rejection sensitivity hasn't been studied as extensively in other personality disorders or in non-clinical populations, leaving a gap in our understanding of how these constructs might interplay. Could an ingrained sensitivity to rejection contribute to reduced pro-social behavior such as a failure to show trustworthiness?

If indeed RS contributes to interpersonal dysfunction through failure to show trustworthiness, could exposing those who exhibit it to cooperative and forgiving interaction partners lead to improved interpersonal outcomes? Research in the fields of behavioral economics and psychology has explored how positive social interactions influence cooperation and pro-social behavior. The use of the repeated trust game (RTG), a well-established experimental approach, has allowed for the analysis of the development of trust through ongoing interactions [@joyce_trust_1995]. In this paradigm, cycles of mutual trust, where each party's trust is reciprocated with trustworthiness, have the effect of enhancing cooperative behaviors even among individuals who are initially inclined to be distrustful [@king-casas_getting_2005]. @fowler_cooperative_2010 studied behavior in social networks interacting in a public goods game and found that cooperative behavior tends to cluster, suggesting that exposure to cooperative peers can lead to more cooperative behavior. Similarly, research on social learning theory [@bandura_social_1977] has long demonstrated that individuals learn and model the behavior of those around them, indicating that if someone is consistently exposed to cooperative and positive individuals, they're likely to emulate this behavior. These insights highlight that engaging with compassionate and forgiving others can be an effective method for mitigating deeply ingrained mistrust and potentially fostering trustworthy behavior.

In this study, we use a randomized controlled online experiment to test whether exposing participants with varying RS levels to forgiving and more cooperative co-players results in more trustworthy behavior and a repair of potential breakdowns in RTG cooperation. To simulate realistic social interaction while maintaining a high degree of experimental control, we take a novel paradigmatic approach: We use generative models of how humans play the RTG to design an agent that plays the role of the investor, based on Hidden Markov Models (HMMs) fitted to real players' data. A key aspect of these agents is that their actions depend on a latent "trust state" which reacts dynamically to the trustees' returns, simulating real-life trust-building scenarios. An advantage of having such a generative model of behavior is the possibility of controlling different aspects of the agent's strategy such as its general policy, the propensity to cooperate actively, or the propensity to retaliate after breakdowns of cooperation. To further mimic real-world interactions and examine participants' responses to one-off breakdowns of cooperation, we incorporate occasional pre-programmed low investments by the agent.

We pre-screened participants for high or low rejection sensitivity (RS) using a validated questionnaire, then assigned them exclusively to the trustee role in a series of trust games. After playing a 15-round RTG with a human-like HMM investor, they were randomly assigned to either a Control or Manipulation condition. In the Manipulation condition, participants were exposed over three RTGs to HMM investors designed with a limited propensity for retaliation, potentially mitigating ingrained mistrust. In the Control condition, participants played three RTGs against the same human-like HMM. After this exposure phase, all participants played another 15-round RTG with a human-like HMM investor, similar to the one in the pre-exposure phase. We expect those exhibiting high RS to be more responsive to the pre-programmed change in the agent's cooperativeness and retaliation propensity. We also hypothesize that those in the Manipulation condition would behave more cooperatively and have a lower propensity to retaliate to the co-player's defection after the exposure phase, with those in the higher RS group showing a bigger effect.

@@

To have participants with large differences in RS, a total of 1195 participants were pre-screened on the Prolific Academic platform (prolific.co) using the Rejection Sensitivity Questionnaire (RSQ) to finally select two similarly sized groups: One with high RS (RSQ score > 15) and the other with low RS (RSQ score < 10) totalling 206 participants (56% female). These were then invited through prolific to take part in the main experiment. The required sample size was determined using an *a priori* power analysis to have an 80% probability to detect a small effect size with a 5% type I error rate in a repeated measures ANOVA with between factors. The mean age of participants was 34.6 years, with an 11.9 years standard deviation. Participants were paid a fixed fee of £6 plus a bonus payment dependent on their performance that averaged £0.5.

@@

The experiment had a 2 (Condition: Manipulation or Control) by 2 (RS : High or Low) by 2 (Phase: Trust-Game Pre Manipulation, Trust-Game Post Manipulation) design, with repeated measures on the Phase factor (Figure \@ref(fig:HMMPanels).A). Participants within each pre-screened group were randomly assigned to one of the two levels of the Condition factor. The games were designed and implemented online using Empirica v1 [@almaatouq_empirica_2021]. The planned experiment received approval from the University of Heidelberg's Medical Faculty ethics commission (ID:S-708/2023) and the experiment was performed in accordance with the ethics board guidelines and regulations. All participants provided informed consent prior to their participation.

@@

```{r HMMPanels, fig.cap= "Panel A: a timeline of the experiment for both conditions. The RTG is played in dyads, with participants always assigned the role of the trustee and the HMM agent that of the investor. The investor is endowed with 20 units at the start of each round. They need to decide how much of that endowment they want to invest with the trustee. The investment is then multiplied by a factor of 3 and sent to the trustee who needs to decide how much of the multiplied investment they want to send back to the investor. The difference between conditions is the type of agents participants are exposed to in the exposure phase. Panels B - D: We construct the artificial investor agent by fitting a three-state hidden Markov model to data of human investors engaged in the 10 round RTG. From the fitted HMM, we get the distribution of investments by the human-like agent, conditional on its latent state as shown in Panel B. The fitted HMM also yields the transition probability of the agent to a state on trial t+1 as a function of the net return (difference between the investment sent and the amount received in return) on trial t as shown in Panel C. Each plot in Panel C represents a different starting latent state on trial t, and each line represents the probability of transitioning to a particular state in trial t+1. Panel D shows the transition probabilities of the forgiving HMM agent in the low trust state. Unlike the human-like HMM, the forgiving HMM always transitions out of the low-trust state, and is more likely to end up in a high-trust state, as a proxy for coaxing behavior. Transitions in the medium and high trust states were identical for both agents.", include=T, fig.width=11, fig.height=8.5, out.width="100%", dpi=300, dev='png'}

@@

# Function to scale image while maintaining aspect ratio
scale_image <- function(img, target_width, target_height) {
  img_info <- image_info(img)
  aspect_ratio <- img_info$width / img_info$height

  if (aspect_ratio > (target_width / target_height)) {
    new_width <- target_width
    new_height <- target_width / aspect_ratio
  } else {
    new_height <- target_height
    new_width <- target_height * aspect_ratio
  }

  image_scale(img, paste0(new_width, "x", new_height))
}

# Calculate target dimensions (97% of half the page to leave minimal margin)
page_width <- 11 * 300  # 11 inches at 300 DPI
page_height <- 8.5 * 300  # 8.5 inches at 300 DPI
target_width <- (page_width * 0.97) / 2
target_height <- (page_height * 0.97) / 2

# Scale images
image1 <- scale_image(image1, target_width, target_height)
image2 <- scale_image(image2, target_width, target_height)
image3 <- scale_image(image3, target_width, target_height)
image4 <- scale_image(image4, target_width, target_height)

# Annotate each image with a label
label_size <- 72
image1 <- image_annotate(image1, text = "A", location = "+10+10", size = label_size, color = "black", font = "Arial-Bold")
image2 <- image_annotate(image2, text = "B", location = "+10+10", size = label_size, color = "black", font = "Arial-Bold")
image3 <- image_annotate(image3, text = "C", location = "+10+10", size = label_size, color = "black", font = "Arial-Bold")
image4 <- image_annotate(image4, text = "D", location = "+10+10", size = label_size, color = "black", font = "Arial-Bold")

# Add a small border to each image to create spacing
border_size <- 5
image1 <- image_border(image1, "white", paste0(border_size, "x", border_size))
image2 <- image_border(image2, "white", paste0(border_size, "x", border_size))
image3 <- image_border(image3, "white", paste0(border_size, "x", border_size))
image4 <- image_border(image4, "white", paste0(border_size, "x", border_size))

# Create the montage
combined_image <- image_append(
  c(image_append(c(image1, image2), stack = TRUE),
    image_append(c(image3, image4), stack = TRUE))
)

# Display the combined image
plot(combined_image)
```

@@

@@

<!-- EFFECT SIZES -->
```{r}
# Load the necessary packages

@@

```
To examine this four-way interaction, we conducted a contrast analysis of how the effect of investment on returns changed from pre- to post-manipulation for different rejection sensitivity (RS) groups in both conditions.
Starting with the Manipulation condition, for participants with low RS, the effect of investment on returns increased significantly from pre- to post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipLowRS`. This suggests that after the manipulation, low RS participants became more responsive to their co-player's investments, returning proportionally more as investments increased.
In contrast, for participants with high RS, the effect of investment on returns decreased significantly from pre- to post-manipulation,
`r papaja::apa_print(results_4w)$full_result$DiffininveffectPost_PreManipHighRS`. This indicates that high RS participants became less responsive to their co-player's investments after the manipulation, with smaller increases in returns as investments increased.
The difference in these pre-post changes between high and low RS groups was significant,
`r papaja::apa_print(results_4w)$full_result$DiffinPost_PrebetweenHighandLowRS`. This result suggests that the manipulation had significantly different effects on how low and high RS participants responded to their co-player's investments.

@@

<!-- Focusing on post defection only:  -->
### Post Defection Trials

@@

### HMM Investor in Pre and Post Phases
<!-- Linear mixed effects model for investments -->
Was the HMM's strategy similar between pre and post phases in the control condition? Was participants' behavior post exposure differentiated enough to induce a different reaction from the HMM? To answer these questions, we test for differences in the HMM agent's investment by Phase, Condition and RS using a linear mixed-effects model as described in the methods section. As seen in Figure \@ref(fig:boxPlots), we find no main or interaction effects, indicating the HMM's behavior was on aggregate similar across levels of Phase, Condition and RS. This consistency in the investor's behavior is a desirable feature of the HMM agent when the participants behavior is largely simialr between phases. More importantly, it indicates that the lower returns of participants in the post phase of the manipulation condition were not differentiated enough to make the HMM react by transitioning to lower latent trust states. It is also noteworthy that the HMM agent was relatively successful in imitating human behavior in this paradigm: When asked during debrief whether they thought the investors they faced were human or not, $41$% of participants thought they were either facing a human or were not sure of the nature of the co-player. When asked to justify their choice, many answers reflected participants projecting human traits such as "spitefulness" or "greed" onto the artificial co-player's behavior.

@@

emcheck_inv <- pairs(emmeans::emmeans(mod_invs_expo, c("condition.f"), pbkrtest.limit = 6300))
emcheck_inv
```
So far we focused on analysing behavior for the pre and post phases. Here, we look at returns and investments in the exposure phase. The linear mixed effects model of participants' returns in the exposure phase does not show a main effect of Condition on returns. There was a main effect of Investment, `r papaja::apa_print(mod_returns_expo)$full_result$inv_scaled`, with participants positively reciprocating higher investments, an interaction effect between Condition and Investment `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled)`, showing a stronger positive reciprocity in the Control condition, and finally a three way interaction between the RS group, Condition and Investment `r gsub("(\\d{1,3}),(\\d{3})", "\\1\\2",papaja::apa_print(mod_returns_expo)$full_result$condition_f_inv_scaled_high_RS)`, showing that this stronger positive reciprocity to investment in the Control condition is higher for participants with high RS. The linear mixed effects model of the HMM investments shows a main effect of Condition `r papaja::apa_print(mod_invs_expo)$full_result$condition_f`, suggesting higher overall investments for the forgiving HMM  compared to the human-like HMM, but no difference in investments when facing low and high RS groups.

@@

### Comparing Pre and Post Ratings
<!-- Comparing post-exposure to exposure ratings, participants in the Manipulation condition rated the investors they faced in the exposure phase as less cooperative,  -->
<!-- `r papaja::apa_print(comp3_coop)$full_result$Manipulation_PostvsExpo`, less forgiving `r papaja::apa_print(comp3_forg)$full_result$Manipulation_PostvsExpo`, and were less willing to face them again `r papaja::apa_print(comp3_again)$full_result$Manipulation_PostvsExpo`. Ratings for those in the Control group did not differ on how cooperative, forgiving the investors were, and on willingness to face them again.  -->

@@

```
```{r}
# Filter for expo games only
filtered_expo <- d_long %>%
  filter(gameNum.f %in% c("expo1", "expo2", "expo3"))

@@

We used a randomized controlled online experiment where participants played a RTG with artificial agents designed to simulate human-like trust-building scenarios. Participants were then exposed to either forgiving or human-like HMM agents before playing another RTG. Contrary to our hypothesis, exposure to forgiving agents did not increase trust or cooperation. Instead, participants reduced their returns overall whilst the returns of those in the Control group did not change between the pre and post phase of the experiment. Neither did participants show more forgiving behavior in the Manipulation condition after the one-off defection by the agent.  Why did participants reduce their returns even though they were repeatedly exposed to more cooperative and more forgiving artificial agents? A look at how the participants rated their co-players might shed some light on what might be driving this reduction in returns for those in the Manipulation condition.

Those exposed to the forgiving agent rated their opponent in the post-exposure phase lower on all attributes even though they faced the same dynamic human-like HMM as pre-exposure. One possible explanation for this drop in rating is that participants exhibited a negative contrast effect. This occurs when the evaluation of a person, object, or situation is influenced by comparisons with recently encountered contrasting objects or people. If we've recently interacted with someone exceptionally nice, our perception of a normal level of niceness might be skewed, making normal behavior seem less favourable or even negative by comparison [@kobre_negative_1972]. As the most recently faced opponents were more forgiving and cooperative, this negative contrast effect may have trumped any learning transfer from being repeatedly exposed to cooperative and forgiving agents [@zentall_within-trial_2005]. If this contrast effect is indeed replicable, then an avenue for future research would be to use it to our benefit by making the participants play agents with low cooperation perception.
